# Measuring and analysing the causal impact of the Fukushima Daiichi nuclear disaster

```{r sweaveopts}
library(knitr)
opts_chunk$set(comment=NA, fig.width=8, fig.height=6,cache=FALSE,self.contained=FALSE)
```

```{r}
library(rdd)
library(ggplot2)
library(foreign)
library(AER)
```


## Data

We use the International Social Survey Program (ISSP) data 2010 which added an environmental attitudes survey as a survey focus. The survey was conducted between 2010 and 2011 whereby the exact period varies by country. Data needs to be downloaded from ZACAT <http://zacat.gesis.org/webview/>.

```{r dataimport}
library(foreign)
try(setwd("C:/Users/Hackstutz/Dropbox/Fukushima/"))
try(setwd("C:/Users/rudi/Dropbox/Fukushima/"))
issp <- read.dta("issp_data/ZA5500_v2-0-0.dta")
```

For a few countries we have information about the month the interview took place but not the exact day. We suggest to impute these information using random uniform distributions.

```{r datapreperation}
#random.org
set.seed(950075)

issp$randomDY <- NULL
howmany31 <- length(which(issp$DATEMO%in%c("January","March","May","July","August","October","December")))
howmany30 <- length(which(issp$DATEMO%in%c("April","June","September","November")))
howmany28 <- length(which(issp$DATEMO%in%c("February")))
# Plus cases missing DATEMO Sums to N (45199)

# Create random days that can be used to fill Missings
issp$randomDY[issp$DATEMO%in%c("January","March","May","July","August","October","December")]<-floor(runif(howmany31,1,32))
issp$randomDY[issp$DATEMO%in%c("April","June","September","November")]<-floor(runif(howmany30,1,31))
issp$randomDY[issp$DATEMO%in%c("February")]<-floor(runif(howmany28,1,29))

# Fill where the value of DATEDY is missing:
issp$DATEDY[is.na(issp$DATEDY)]<-issp$randomDY[is.na(issp$DATEDY)]

# Create a nice data variable
levels(issp$DATEYR) <- c("2009","2010","2011","No answer")
issp$datestring <- paste(issp$DATEYR,issp$DATEMO,issp$DATEDY,sep="-")
# Change the locale to apply strptime with english names for months
Sys.setlocale("LC_TIME","us")
issp$date <- strptime(issp$datestring,format="%Y-%B-%d")
#issp$V4[is.na(issp$date)] #6471 Missings
                              
# dependent variable
issp$nuclear <- 5-as.numeric(issp$V45)
issp$nuclear[issp$nuclear%in%c("Can't choose","NA")] <- NA
issp$nuclear <- factor(issp$nuclear)

# independent variable: before/after fukushima
fukushima <- strptime("2011-03-13",format="%Y-%m-%d")
issp$after <- issp$date >= fukushima

# controls
issp$country <- issp$V4
issp$sex <- factor(issp$SEX)
issp$age <- issp$AGE/10
issp$old <- issp$age>4.7
# calculate standardized household income by country
issp$hhincome <- NA
for(i in which(names(issp)=="AT_INC"):which(names(issp)=="ZA_INC")) {
  issp$hhincome[!is.na(issp[,i])] <- scale(as.numeric(issp[!is.na(issp[,i]),i]))
}

issp$welloff <- issp$hhincome > 0
issp$educated <- issp$DEGREE%in%c("University degree completed, BG: lower tertiary level, upper tertiary level","University degree incomplete (qualifications above higher secondary level)","Higher secondary completed (usual entry requirements for universities)")
issp$trust <- (as.numeric(issp$V11)+as.numeric(issp$V12))/2
issp$pol_trust <-5-as.numeric(issp$V13)

```

It is important to know which countries contribute information to which period of time.

```{r timelines}
library(ggplot2)

ggplot(issp, aes(x=date,y=V4))+geom_point()+geom_vline(xintercept=as.numeric(fukushima),color="gray")+theme_bw()
```

## Method

### Regression Discontinuity Design (RDD)

RDD's causal identification strategy is to compare cases which are very close to a certain cutoff point of threshold which assigns data into a control and treatment group. Concerning the Fukushima accident, the assignment variable is time and the threshold is the day of the accident: 13th March 2011. Whether the interview was conducted before the 13th or after appears to be random so comparing individuals surveyed just before and just after the accident does control for most unobserved heterogeneity. RDD is illustrated in the following figure:

```{r rddillustration}
library(ggplot2)
library(grid) #for function arrow()
x<-1:100
set.seed(123)
y<-rnorm(100)
y[50:100]<-y[50:100]+2
group<-c(rep(0,50),rep(1,50))
qplot(x,y,group=group) + stat_smooth(se=FALSE)+geom_text(show_guide=FALSE,x=50,y=1.2,label="Treatment \neffect",color="black")+geom_segment(aes(x = 50, y = 1.5, xend = 50, yend = 2.2), arrow = arrow(),color="black")+geom_segment(aes(x = 50, y = 0.8, xend = 50, yend = 0.45), arrow = arrow(),color="black")+ggtitle("RDD Visualization")+theme_bw()
```

### Difference-in-Difference Estimator (DiD)

DiD's causal identification strategy is to get rid of multiple sources of unobserved heterogeneity by controlling for country-characteristics (country dummies) and period-characteristics (period dummies) which in our case is accomplished by estimating a Country-Fixed-Effects Modell including a "before/after-Fukushima-Dummy". There are two key assumptions to mention. First, DiD assumes the timetrend of $y$, i.e. attitudes towards nuclear energy to be stable over time. Second, as there are countries which are completely before or completely after the accident, the covariance of all variables needs to be stable accross countries. (das müssen wir noch diskutieren). DiD is illustrated in the next figure:

```{r didillustration}
time <- rep(c(0,1),3)
observed <- c(30,15)
control <- c(20,35)
counterfactual <- c(30,45)
group <- factor(c(1,1,2,2,3,3), labels=c("observed","control","counterfactual"))
y <- c(observed, control, counterfactual)
qplot(time,y,group=group, colour=group,ylim=c(0,50),xlim=c(0,1.06))+geom_line(size=1)+ggtitle("DiD Visualization")+geom_segment(aes(x = 1, y = 27, xend = 1, yend = 15.5), arrow = arrow(),color="black")+geom_segment(aes(x = 1, y = 33, xend = 1, yend = 44.5), arrow = arrow(),color="black")+geom_text(show_guide=FALSE,x=1,y=30,label="Treatment \neffect",color="black")+theme_bw()
```


## Models

### RDD

```{r parametric}
issp$t <- as.numeric(issp$date)-as.numeric(fukushima)
# Exlude imputated Canadians:
issp$valid <- TRUE
issp$valid[issp$country=="CA-Canada"& issp$DATEMO=="March"] <- FALSE

fit_rd_param<-lm(as.numeric(nuclear)~after*(sex+old+educated)*(t+I(t^2)+I(t^3)+I(t^4)+I(t^5)+I(t^6)), data=subset(issp,valid==TRUE))
# interaction is significant for sex but not for old or educated
rd_data <- model.frame(fit_rd_param)
rd_data$pred <- predict(fit_rd_param)
ggplot(rd_data, aes(y=pred,x=t))+geom_point()+facet_wrap(~sex)
ggplot(rd_data, aes(y=pred,x=t))+geom_point()+facet_wrap(~old)
ggplot(rd_data, aes(y=pred,x=t))+geom_point()+facet_wrap(~educated)
```


```{r rdd}
library(rdd)
# Efficiency gains can be made by including covariates. I.e. country dummies. Austin Nicols stronly discourages that.
#RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=issp)
bw <- 86400*c(14,30,60,90) #86400 equals 1 day



fit_rd<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,valid==TRUE),bw=bw[2])

par(mfrow=c(1,1))
plot(fit_rd)

# Split by Sex
fit_rd_male<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,sex=="Male"&valid==TRUE),bw=bw[2])
fit_rd_female<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,sex=="Female"&valid==TRUE),bw=bw[2])

par(mfrow=c(1,2))
plot(fit_rd_male)
title("Effect for Males")
plot(fit_rd_female)
title("Effect for Females")

# Split by Age (old means >47 years)
fit_rd_young<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,old==FALSE&valid==TRUE),bw=bw[2])
fit_rd_old<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,old==TRUE&valid==TRUE),bw=bw[2])

par(mfrow=c(1,2))
plot(fit_rd_young)
title("Effect for young people")
plot(fit_rd_old)
title("Effect for old people (>47y)")


# Split by income (above/below median) 
fit_rd_welloff<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,welloff==TRUE&valid==TRUE),bw=bw[2])
fit_rd_notwell<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,welloff==FALSE&valid==TRUE),bw=bw[2])

par(mfrow=c(1,2))
plot(fit_rd_welloff)
title("hh income above median")
plot(fit_rd_notwell)
title("hh income below median")

# Split by education (qualified to enter university and higher)
fit_rd_uneduc<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,educated==FALSE&valid==TRUE),bw=bw[2])
fit_rd_educ<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,educated==TRUE&valid==TRUE),bw=bw[2])

par(mfrow=c(1,2))
plot(fit_rd_uneduc,ylim=c(3,5))
title("Effect for less educated")
plot(fit_rd_educ,ylim=c(3,5))
title("Effect for more educated")

# Table with all effects
betas30 <- c(fit_rd$est[1], fit_rd_male$est[1],fit_rd_female$est[1],fit_rd_young$est[1],fit_rd_old$est[1],fit_rd_uneduc$est[1],fit_rd_educ$est[1],fit_rd_welloff$est[1],fit_rd_notwell$est[1])
betas60 <- c(fit_rd$est[3], fit_rd_male$est[3],fit_rd_female$est[3],fit_rd_young$est[3],fit_rd_old$est[3],fit_rd_uneduc$est[3],fit_rd_educ$est[3],fit_rd_welloff$est[3],fit_rd_notwell$est[3])
coefnames <- c("Overall", "Males","Females","Young","Old","Less educated","More educated", "Above median household income","Below median household income")
se30 <-  c(fit_rd$se[1], fit_rd_male$se[1],fit_rd_female$se[1],fit_rd_young$se[1],fit_rd_old$se[1],fit_rd_uneduc$se[1],fit_rd_educ$se[1],fit_rd_welloff$se[1],fit_rd_notwell$se[1])
se60 <-  c(fit_rd$se[3], fit_rd_male$se[3],fit_rd_female$se[3],fit_rd_young$se[3],fit_rd_old$se[3],fit_rd_uneduc$se[3],fit_rd_educ$se[3],fit_rd_welloff$se[3],fit_rd_notwell$se[3])

order <- 9:1
min30 <- betas30-1.96*se30
max30 <- betas30+1.96*se30
min60 <- betas60-1.96*se30
max60 <- betas60+1.96*se30

res_data <- data.frame(coefnames,betas30,betas60,min30,max30,order,min60,max60)

ggplot(res_data, aes(y = betas30, ymin=min30, ymax=max30, x = reorder(coefnames,order))) + geom_pointrange() + coord_flip()+theme_bw()+scale_x_discrete("Effects")+scale_y_continuous("beta") + geom_hline(xintercept = 0, colour="red")

ggplot(res_data, aes(y = betas60, ymin=min60, ymax=max60, x = reorder(coefnames,order))) + geom_pointrange() + coord_flip()+theme_bw()+scale_x_discrete("Effects")+scale_y_continuous("beta") + geom_hline(xintercept = 0, colour="red")

```

### DiD

Question: do we need ordered logit here because of potential compression effects (i.e. groups that had radical opinions BEFORE fukushima are not able to change there attitude. Image all women answered "very dangerous" before fukushima. then we can not observe an effect afterwards). 

We can answer that with a simulation using plausible numbers (obtained from the later analyses)

```{r}
library(MASS)
woman <- rbinom(36342,1,0.5)
# introduce an effect for sex (-0.33) and after (-0.21) but no interaction
# also testet for the effects if nothing else is controlled for: sex=-0.348, after=-0.379
#b_sex <- -0.33
#b_after <- -0.21
b_sex <- -0.348
b_after <- -0.379

y_star <- woman*b_sex+rnorm(36342)
#retrieve cut-off-points
q20<-quantile(y_star,0.2)
q40<-quantile(y_star,0.4)
q60<-quantile(y_star,0.6)
q80<-quantile(y_star,0.8)
# introduce the fukushima effect
after <- rbinom(36342,1,0.5)
y_star <- y_star+b_after*after
y <- ifelse(y_star<q20,1,ifelse(y_star<q40,2,ifelse(y_star<q60,3,ifelse(y_star<q80,4,5))))
summary(lm(y~woman*after))
summary(polr(as.factor(y)~woman*after))

```

We see a bias if we use -2 for women (so there must be an extreme base line difference for this artefact to occure). ologit solves the problem but there is no need to use it. For our FE model it is better to stick to metric assumptions.


```{r did}
# We need ordered logit regression here. Fixed effects can be achieved by including country-dummies.
library(MASS)
# including country dummies in polr() throws an error.. 
#summary(polr(nuclear~after*(sex+age), data=issp))

# using OLS vs ordered logit seems legit (see simulation)

# Model without interactions to get the gross effect
summary(lm(as.numeric(nuclear)~after+sex+age+DEGREE+trust+pol_trust+as.numeric(TOPBOT)+country+DATEMO*DATEYR, data=issp))
# the overall effect is 0.15 on the 1-5 scale. 

fit <- lm(as.numeric(nuclear)~after*(sex+age+DEGREE+trust+pol_trust+as.numeric(TOPBOT))+country+DATEMO*DATEYR, data=issp)
summary(fit)
# controlling for country trends does not affect our results
fit2<-lm(as.numeric(nuclear)~after*(sex+age+DEGREE+trust+pol_trust)+country+DATEMO*DATEYR+country:as.numeric(date), data=issp)
summary(fit2)

# only use countries that include both: before and after measure
fit3<-lm(as.numeric(nuclear)~after*(sex+age+DEGREE+trust+pol_trust)+country+DATEMO*DATEYR+country:as.numeric(date), data=subset(issp,country%in%c("CH-Switzerland","SI-Slovenia","DK-Denmark","CA-Canada")))
summary(fit3)

# test social desirability
summary(lm(as.numeric(nuclear)~as.numeric(after)*MODE+country+DATEMO*DATEYR+country:as.numeric(date), data=issp))

library(survey)# contains regTermTest for F-test

regTermTest(fit,"after:DEGREE") 
regTermTest(fit2,"after:DEGREE")#no interaction with education

# to present the results it is easier to show two interaction effects instead of main effect plus interaction:
issp$before <- issp$after==FALSE
fit_final<-lm(as.numeric(nuclear)~as.numeric(before):(as.numeric(sex)+age+DEGREE+trust+pol_trust+hhincome)+as.numeric(after):(as.numeric(sex)+age+DEGREE+trust+pol_trust+hhincome)+country+DATEMO*DATEYR+country:as.numeric(date), data=issp)

# test interactions
library(AER)
coeftest(lm(as.numeric(nuclear)~as.numeric(after)*(as.numeric(sex)+age+DEGREE+trust+pol_trust+hhincome)+country+DATEMO*DATEYR+country:as.numeric(date), data=issp))[1:2,]

#test endogeneity of trust
bw <- 86400*c(14,30,60,90) #86400 equals 1 day
fit_rd_trust<-RDestimate(as.numeric(trust)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,valid==TRUE),bw=bw[2])
fit_rd_poltrust<-RDestimate(as.numeric(pol_trust)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,valid==TRUE),bw=bw[2])
par(mfrow=c(1,2))
plot(fit_rd_trust)
title("Impact on Trust")
plot(fit_rd_poltrust)
title("political Trust")



# robustness check: shifting 10 days left and right
fit_rd_left<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima-10*3600*24),data=subset(issp,valid==TRUE),bw=bw[2])
fit_rd_right<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima+10*3600*24),data=subset(issp,valid==TRUE),bw=bw[2])
fit_rd<-RDestimate(as.numeric(nuclear)~as.numeric(date), cutpoint=as.numeric(fukushima),data=subset(issp,valid==TRUE),bw=bw[2])
par(mfrow=c(1,3))
plot(fit_rd_left)
title("-10 days")
plot(fit_rd)
title("March 13th")
plot(fit_rd_right)
title("+10 days")

# because of our strong assumptions (metric Y) we should use robust standard errors
results<-coeftest(fit_final, vcov=vcovHC(fit_final,type="HC0"))
results

```

### Plotting the results

We controlled for country levels, period dummies (monthly) and linear time trends within each country


```{r airplane_plot}
coefnames <- c("Female","Age (10y)","Lowest formal qualification","Intermediate secondary qualif.","Higher secondary","University incomplete","University","General Trust","Political Trust", "Household income")
maineffects <- results[c(40:49),1]
interactions <- results[c(50:59),1]
main_se <- results[c(40:49),2]
int_se <- results[c(50:59),2]
order <- 10:1
main_min <- maineffects-1.96*main_se
main_max <- maineffects+1.96*main_se
int_min <- interactions-1.96*int_se
int_max <- interactions+1.96*int_se

res_data <- data.frame(coefnames,maineffects,interactions,main_se,int_se,order,main_min,main_max,int_min,int_max)

ggplot(res_data, aes(y = maineffects, ymin=main_min, ymax=main_max, x = reorder(coefnames,order))) + geom_pointrange() + coord_flip()+theme_bw()+scale_x_discrete("Effects")+scale_y_continuous("beta") + geom_hline(xintercept = 0, colour="red") + geom_pointrange(aes(y=interactions,ymin=int_min,ymax=int_max,x=as.numeric(reorder(coefnames,order))-0.1),colour="#00AA00") + annotate("text", x = 5.3, y = 0.25, label = "Before", colour = "black", size=4) + annotate("text", x = 4.6, y = 0.242, label = "After", colour = "#00AA00", size=4)
```

